---
title: "Analytical report"
format: 
  html:
    toc: true
    theme: default
---

### Introduction

This webpage contains the description and codes for reproducing the analysis conducted in the article "Comparative susceptibility of *Ficus carica* and *Morus nigra* to *Cerotelium fici* isolates in cross-inoculation studies".

### Methodology

Each "treatment" in our analysis was defined as a unique combination of host-pathogen, specifically CFF × F. carica, CFM × F. carica, CFF × M. nigra, and CFM × M. nigra. These treatments were evaluated in four independent experiments, with no randomization of treatments within a single experiment. Each experiment was repeated twice for each host-pathogen combination, resulting in a total of eight trials (two repetitions per treatment at different times).

Given the lack of randomization within experiments and the independent nature of the trials, a meta-analytic approach was employed to analyze the data. This approach allowed us to estimate the effects of treatments (treatment means) for each variable of interest while accounting for both within-experiment variability (variance among replicates within the same experiment) and between-experiment variability (differences across independent experiments). By using this method, we were able to pool data across experiments to provide robust and generalizable estimates of treatment effects, while properly addressing the hierarchical structure of the data. The meta-analysis was conducted using a multilevel random-effects model implemented in the metafor package in R.

The model incorporated a fixed effect for treatments and a random effect for experiments to account for variability between experiments. The model's fit was evaluated using restricted maximum likelihood (REML) estimation. Pairwise comparisons between treatments were conducted using the emmeans package to calculate marginal means and perform post hoc tests, adjusted for multiple comparisons using the Tukey method. Estimated marginal means for each treatment were calculated to assess differences in the mean values. Pairwise comparisons were performed to identify statistically significant differences between treatments.

### Load the data

```{r}
#| warning: false
#| message: false
library(readxl)
library(dplyr)
```

### MA for severity

```{r}
#| warning: false
#| message: false

# Load data
datasev <- read_excel ("data_experiments.xlsx")

# Summarize data
summary_data <- datasev %>%
  group_by(interacao, exp) %>%                      
  summarize(
    mean_sev = mean(sev, na.rm = TRUE),  
    var_sev = var(sev, na.rm = TRUE),    
    n = n()                                         
  )
summary_data

summary_data <- summary_data %>%
  mutate(
    SE_sev = sqrt(var_sev / n),  # SE = sqrt(variance / sample size)
    vi = SE_sev^2                      # Variance = SE^2
  )
summary_data <- summary_data |> 
  filter(interacao != "CFFM. nigra")

library(metafor)

# Meta-regression using the summarized data
res <- rma.mv(
  yi = mean_sev,  
  V = vi,              
  mods = ~ interacao, 
  random = ~ interacao | exp, 
  data = summary_data,
  method = "REML"
)
# Summary of results
summary(res)

library(emmeans)
prep_emmeans <- emmprep(res,  mods = ~ interacao)

# Marginal means for treatments
emmeans_res <- emmeans(prep_emmeans, ~ interacao)

# Pairwise comparisons
pairs(emmeans_res)

# Marginal means for interaction by experiment
emmeans_res_exp <- emmeans(prep_emmeans, ~ interacao)

emmeans_res_exp
library(multcomp)
cld(emmeans_res_exp)

```

### MA for number of lesions

```{r}
#| warning: false
#| message: false

# Load data
datalesao <- read_excel("data_experiments.xlsx", "lesao")

# Summarize data
summary_data <- datalesao %>%
  group_by(interacao, exp) %>%                    
  summarize(
    mean_lesoes = mean(lesoes, na.rm = TRUE),  
    var_lesoes = var(lesoes, na.rm = TRUE),   
    n = n()                                          
  )
summary_data

summary_data <- summary_data %>%
  mutate(
    SE_lesoes = sqrt(var_lesoes / n),  
    vi = SE_lesoes^2                      
  )
summary_data <- summary_data |> 
  filter(interacao != "CFFM. nigra")

library(metafor)

# Meta-regression using the summarized data
res <- rma.mv(
  yi = mean_lesoes,  
  V = vi,              
  mods = ~ interacao, 
  random = ~ interacao | exp, 
  data = summary_data,
  method = "REML"
)
# Summary of results
summary(res)

library(emmeans)
prep_emmeans <- emmprep(res,  mods = ~ interacao)

# Marginal means for treatments
emmeans_res <- emmeans(prep_emmeans, ~ interacao)

# Pairwise comparisons
pairs(emmeans_res)

# Marginal means for interaction by experiment
emmeans_res_exp <- emmeans(prep_emmeans, ~ interacao)

emmeans_res_exp
library(multcomp)
cld(emmeans_res_exp)

residuals <- residuals(res)

# Calculate the number of degrees of freedom
n_obs <- length(residuals)
n_params <- length(coef(res)) + length(res$sigma2)

# Calculate the mean square error (MSE)
MSE <- sum(residuals^2) / (n_obs - n_params)

# Residual standard deviation
residual_sd <- sqrt(MSE)

# Calculate the mean of the predicted values
mean_y <- mean(predict(res)$pred)

# Calculate the CV in the original scale
cv <- (residual_sd / mean_y) * 100

cv
```

### MA for sporulation

```{r}
#| warning: false
#| message: false

# Load data

dataesp <- read_excel ("data_experiments.xlsx", "esporulação")

# Summarize data
summary_data <- dataesp %>%
  group_by(interacao, exp) %>%                    
  summarize(
    mean_esporos = mean(esporos, na.rm = TRUE),  
    var_esporos = var(esporos, na.rm = TRUE),    
    n = n()                                        
  )
summary_data

summary_data <- summary_data %>%
  mutate(
    SE_esporos = sqrt(var_esporos/n),  
    vi = SE_esporos^2                      
  )

library(metafor)

# Meta-regression using the summarized data
res <- rma.mv(
  yi = mean_esporos,  
  V = vi,               
  mods = ~ interacao, 
  random = ~ interacao | exp,  
  data = summary_data,
  method = "REML"
)
# Summary of results
summary(res)
res$tau2
library(emmeans)
prep_emmeans <- emmprep(res,  mods = ~ interacao)

# Marginal means for treatments
emmeans_res <- emmeans(prep_emmeans, ~ interacao)

# Pairwise comparisons
pairs(emmeans_res)

# Marginal means for interaction by experiment
emmeans_res_exp <- emmeans(prep_emmeans, ~ interacao)

emmeans_res_exp
library(multcomp)
cld(emmeans_res_exp)

residuals <- residuals(res)

# Calculate the number of degrees of freedom
n_obs <- length(residuals)
n_params <- length(coef(res)) + length(res$sigma2)

# Calculate the mean square error (MSE)
MSE <- sum(residuals^2) / (n_obs - n_params)

# Residual standard deviation
residual_sd <- sqrt(MSE)

# Calculate the mean of the predicted values
mean_y <- mean(predict(res)$pred)

# Calculate the CV in the original scale
cv <- (residual_sd / mean_y) * 100

cv


```
